---
title: "Science Foundation Ireland Grant Commitments"
author: "Brittnany Graham"
format: html
editor: visual
execute:
  echo: false
---

## Install Packages and upload data

```{r}
# Using R
# Option 1: tidytuesdayR R package 
# install.packages("tidytuesdayR")
# install.packages("forcats")
# install.packages("scales")
library(tidytuesdayR)
library(tidyverse)
library(forcats)
library(scales)


# Using R
# Option 1: tidytuesdayR R package 
## install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2025-03-18')
palmtrees <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-18/palmtrees.csv')

```

## Exploring the data

```{r}

palm <- palmtrees

```

# How We Can Structure This Going Forward

For Tidy Tuesday, we can use this structure:

1.  🔎 Explore data (glimpse, summary)

2.  🧹 Clean data (mutate, filter)

3.  📊 Summarize data

4.  🎨 Visualize

5.  ✨ Refine aesthetics

6.  📄 Render report in Quarto

## What Would Impress in a Tidy Tuesday Submission?

::: callout-note
## Publication plots rarely show 100+ categories
:::

So just show the top 15-25

```{r}
# Step 1: Count how many times each Genus appears

top_genera <- palm %>%
  count(acc_genus) %>%
    # count() groups by Genus and calculates the number of rows for each.
  # It automatically creates a new column called 'n' that stores counts.
  
  mutate(percent = n / sum(n)) %>%
   # Create a new column called 'percent'
  # n / sum(n) calculates the proportion each genus represents
  # relative to the total number of observations
  
  arrange(desc(n)) %>%
  # Sort from largest count to smallest
  
  slice_head(n = 20) %>%
  # Keep only the top 20 most abundant genera
  
  mutate(acc_genus = fct_reorder(acc_genus, n))
# Convert Genus to a factor and reorder levels
  # This ensures bars are ordered by count in the plot

```

```{r}
# Step 2: Create the plot

plot <- ggplot(top_genera, aes(x = n, y = acc_genus)) +
  # ggplot() initializes the plot
  # aes() defines aesthetics:
  #   x = counts
  #   y = Genus names
  
  geom_col(fill = "#1B4F72", width = 0.8) +
  # geom_col() creates bars using precomputed counts (n)
  # fill sets the bar color
  # width controls bar thickness
  
  geom_text(
    aes(label = percent(percent, accuracy = 0.1)),
    # label displays formatted percent values on each bar
    # percent() from scales package converts decimals to percentages
    # accuracy = 0.1 rounds to one decimal place
    
    hjust = -0.1, 
    # hjust controls horizontal positioning of text
    # negative value pushes text slightly outside the bar
    
    size = 3.5
    # controls text size
  ) +
  
  labs(
    title = "Top 20 Genera by Relative Abundance",
    subtitle = "Percent of total observations",
    x = "Count",
    y = NULL
    # y = NULL removes redundant axis label
  ) +
  
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.1))
    # Adds 10% extra space to the right of bars
    # This prevents percent labels from being cut off
  ) +
  
  theme_minimal(base_size = 14) +
  # Clean, professional theme
  # base_size controls overall text size
  
  theme(
    plot.title = element_text(face = "bold"),
    # Makes title bold
    
    panel.grid.major.y = element_blank(),
    # Removes horizontal gridlines (reduces clutter)
    
    panel.grid.minor = element_blank()
    # Removes minor gridlines
  )

plot
```

## Creating a Pareto Chart

Answers the question:

**"How much of the total is explained by the top categories?"**

```{r}
pareto_data <- palm %>%
  
  count(acc_genus) %>%  
  # Count number of observations per genus
  
  arrange(desc(n)) %>%  
  # Sort from largest to smallest
  
  slice_head(n = 20) %>%
  #keeps only the top 20
  
  mutate(
    cumulative_count = cumsum(n),
    # cumsum() creates a running total down the column
    
    cumulative_percent = cumulative_count / sum(n),
    # Divide running total by grand total to get cumulative proportion
    
    acc_genus = fct_reorder(acc_genus, n)
    # Reorder factor levels by count (important for plotting order)
  )
```

### Building the Pareto Plot

```{r}
ggplot(pareto_data, aes(x = acc_genus)) +
  
  # Bar layer (counts) ----------------------------------------------------
  
  geom_col(aes(y = n),
           fill = "#2C7FB8",
           width = 0.8) +
  
  
  # Line layer (cumulative %) --------------------------------------------
  
  geom_line(aes(y = cumulative_percent * max(n), group = 1),
            color = "darkred",
            size = 1) +
  
  # Points on the cumulative line ----------------------------------------
  
  geom_point(aes(y = cumulative_percent * max(n)),
             color = "darkred",
             size = 2) +
  
  
  # Secondary axis --------------------------------------------------------
  
  scale_y_continuous(
    
    name = "Count",
    
    sec.axis = sec_axis(
      ~ . / max(pareto_data$n),
      # This rescales the primary axis back to proportion scale
      name = "Cumulative Percentage",
      labels = percent
    )
  ) +
  
  
  # Flip coordinates for readability -------------------------------------
  
  coord_flip() +
  
  
  labs(
    title = "Pareto Chart of Genera Abundance",
    subtitle = "Bars show counts; line shows cumulative percentage",
    x = NULL
  ) +
  
  
  theme_minimal(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```
