---
title: "Weekly Tidy Tuesday Strategy"
author: "Brittnany Graham"
format: html
editor: visual
---

# Weekly Tidy Tuesday Strategy

1.  Read data dictionary
2.  Ask: What story can I tell?
3.  Pick ***ONE*** clear insight
4.  Build 1 - 2 clean plots
5.  polish typography
6.  Export high resolution

That's it, don't overcomplicate.

### What I want to do next:

Next-level upgrades:

- Parameterized reports (choose week automatically)

- Custom CSS styling

- Automatic weekly folder creation

- GitHub auto-publish

- Interactive plotly version

- Adding narrative commentary blocks

**You are officially moving into:**

> Data scientist who communicates.

If you want, next I can:

- Build you a “Graduate-level Tidy Tuesday workflow"

- ~~Show you how to design plots like The Economist~~

- Help you create a personal ggplot theme you reuse every week

## Midnight Narrative - WINNER

```{r}
midnight_narrative <- c(   
  midnight = "#1D3557",   
  coral = "#E76F51",   
  blush = "#F4A261",   
  rosewood = "#B56576",   
  sage = "#6C9A8B",   
  linen = "#F1FAEE",   
  ink = "#222222" )  
## other colors 
#Darkened base  
"#14213D"  
# Deepen the coral 
"#D8573C" 
```

### 🌌 MINI BRANDING GUIDE

**Visual Hierarchy Rules (Very Important)**

#### Foundation (70% of design)

- Background –\> linen

- Text –\> ink

- Titles –\> midnight

- Most bars –\> neutral/linen or soft grey

#### Highlight (20%)

- Primary highlight –\> coral

- Secondary comparison –\> sage

#### Emphasis (10%)

- Key insight –\> blush

- Annotation accents –\> rosewood

Never use all accent colors at once

Restriction = sophistication

## Palette

```{r}
midnight_narrative <- c(
  midnight = "#1D3557",
  coral = "#E76F51",
  blush = "#F4A261",
  rosewood = "#B56576",
  sage = "#6C9A8B",
  linen = "#F1FAEE",
  ink = "#222222"
)
```

## Reusable Theme Function

```{r}
theme_midnight_narrative <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.background = element_rect(fill = "#F1FAEE", color = NA),
      panel.background = element_rect(fill = "#F1FAEE", color = NA),
      plot.title = element_text(
        face = "bold",
        size = 16,
        color = "#1D3557",
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(
        color = "#222222",
        margin = margin(b = 15)
      ),
      axis.text = element_text(color = "#222222"),
      panel.grid.major.y = element_line(color = "grey85"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "none"
    )
}
```
