---
title: "US Droughts"
author: "Brittnany Graham"
format: 
  html:
    theme: cosmo
    css: midnight-style.css
    code-fold: true
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

## Background

This dataset was covered by the NY Times and CNN.

The dataset ranges from 2001 to 2021.

```{r}

## Installing tidytuesdayR package and downloading the data

# install.packages("tidytuesdayR")
library(tidytuesdayR)
library(tidyverse)
library(janitor)

drought <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-07-20/drought.csv')

```

## Cleaning Data

```{r}
## getting a feel for the data
glimpse(drought)

## Clean column names and fix the dates

drought_clean <- drought %>%
  clean_names() %>%
  mutate(
    map_date = ymd(map_date) #converting the dates - be sure to pay attention to the structure of the date
  )
glimpse(drought_clean)


  ## Always look and see what the dataset looks like before you start running the code so this way you can accurately determine what's happening and reduce the amount of headaches

## Clean and Format drought levels

drought_clean <- drought_clean %>%
  mutate(
    drought_lvl = factor(
      drought_lvl,
      levels = c("None", "D0", "D1", "D2", "D3", "D4"),
      labels = c(
        "No Drought",
        "Abnormally Dry",
        "Moderate Drought",
        "Severe Drought",
        "Extreme Drought",
        "Exceptional Drought"
      )
    )
  )

glimpse(drought_clean)
```

## 📍Filter to Tennessee and summarize

Because this is where I live.

```{r}

# Filtering down to Tennessee
tn_drought <- drought_clean %>%
  filter(state_abb == "TN")

## Summarizing

tn_summary <- tn_drought %>%
  group_by(map_date, drought_lvl) %>%
  summarise(
    area_pct = sum(area_pct, na.rm = TRUE),
    .groups = "drop"
  )
```

### Plot

```{r}

## two color palettes for drought
drought_colors <- c(
  "No Drought" = "#FFFFFF",
  "Abnormally Dry" = "#FCD37F",
  "Moderate Drought" = "#FFAA00",
  "Severe Drought" = "#E60000",
  "Extreme Drought" = "#730000",
  "Exceptional Drought" = "#4B0082"
)

midnight_drought <- c(
  "No Drought" = "#0B132B",
  "Abnormally Dry" = "#1C2541",
  "Moderate Drought" = "#3A506B",
  "Severe Drought" = "#5BC0BE",
  "Extreme Drought" = "#C3423F",
  "Exceptional Drought" = "#6A040F"
)

ggplot(tn_summary, aes(x = map_date, y = area_pct, fill = drought_lvl)) +
  geom_area() +
  labs(
    title = "Drought Conditions in Tennessee Over Time",
    x = "Date",
    y = "Percent Area",
    fill = "Drought Level"
  ) +
  scale_fill_manual(values = drought_colors) +
  theme_minimal()
```

## Land versus population impact 

I want to look at population level effects of drought in Tennessee.
Since Tennessee's land uses about 41% of it's land for agriculture, that translates to approximately 10.8 million acres of agricultural land.
Various crops and livestock contribute significantly to the state's economy.
Drought inevitably impacts the farmers and surrounding population relying on farmers leading to overall lack of food security in the state.

```{r}

# land versus population impact
tn_compare <- tn_drought %>%
  group_by(map_date, drought_lvl) %>%
  summarise(
    area_pct = sum(area_pct, na.rm = TRUE),
    pop_pct = sum(pop_pct, na.rm = TRUE),
    .groups = "drop"
  )

## pivoting for comparison plotting

tn_compare_long <- tn_compare %>%
  pivot_longer(
    cols = c(area_pct, pop_pct),
    names_to = "metric",
    values_to = "percent"
  )

view(tn_compare_long)
```

### Plotting Area vs. Population

```{r}

ggplot(tn_compare_long, aes(x = map_date, y = percent, fill = drought_lvl)) + 
  geom_area() +
  facet_wrap(~ metric, ncol = 1) +
  scale_fill_manual(values = drought_colors) +
  labs(
    title = "Drought Impact: Land vs. Population in Tennessee",
    y = "Percent",
    x = "Date",
    fill = "Drought Level"
  ) +
  theme_minimal()

```

## Showcasing the Tennessee Map 

```{r}

# install.packages("rnaturalearth")
# install.packages("rnaturalearthdata")

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(gganimate)
```

## Get the US State Boundaries

```{r}
us_states <- ne_states(country = "United States of America", returnclass = "sf")

glimpse(us_states)

## will be using "postal" because it matches the "state_abb" from the drought dataset
```

## Prepare drought data for mapping

Picking a single date first (maps are easiest when static):

```{r}

drough_map <- drought_clean %>%
  filter(map_date == as.Date("2021-07-13")) %>%
  group_by(state_abb, drought_lvl) %>%
  summarise(area_pct = sum(area_pct, na.rm = TRUE),
            .groups = "drop")
```

## Join Spatial & Drought Area

```{r}
drought_state <- drought_clean %>%
  filter(map_date == as.Date("2021-07-13")) %>%
  group_by(state_abb, drought_lvl) %>%
  summarise(area_pct = sum(area_pct), .groups = "drop") %>%
  group_by(state_abb) %>%
  slice_max(area_pct, n = 1) %>%
  ungroup()
```

Now join to geometry

```{r}
us_drought_sf <- us_states %>%
  left_join(drought_state, by = c("postal" = "state_abb"))

## Zoomed in on Tennessee

tn_map <- us_drought_sf %>%
  filter(postal == "TN")

ggplot(tn_map) +
  geom_sf(aes(fill = drought_lvl)) +
  scale_fill_manual(values = drought_colors, na.value = "grey80") +
  labs(
    title = "Tennessee Drought Conditions",
    fill = "Drought Level"
  ) +
  theme_minimal()
```

## Creating a animated map over time

This is for better storytelling

```{r}
tn_sf <- us_states %>%
  filter(postal == "TN")
```

**Prepare the Tennessee drought time series**

```{r}

tn_drought_ts <- drought_clean %>%
  filter(state_abb == "TN") %>%
  group_by(map_date, drought_lvl) %>%
  summarise(area_pct, sum(area_pct, na.rm = TRUE), .groups = "drop")
```

**Join TN geometry to each time step**

```{r}
tn_anim_sf <- tn_sf %>%
  slice(rep(1:n(), each = n_distinct(tn_drought_ts$map_date))) %>%
  mutate(map_date = rep(sort(unique(tn_drought_ts$map_date)),
                        each = nrow(tn_sf)))

## now join drought info:

tn_anim_sf <- tn_anim_sf %>%
  left_join(tn_drought_ts, by = 'map_date')
```

### Building the base plot

```{r}
p <- ggplot(tn_anim_sf) +
  geom_sf(aes(fill = drought_lvl), color = "white", linewidth = 0.5) +
  scale_fill_manual(values = drought_colors, na.value = "grey80") +
  labs(
    title = "Tennessee Drought Conditions over 10 Years",
    subtitle = "Date: {frame_time}",
    fill = "Drought Level"
  ) +
  theme_minimal()

## animate over time

anim <- p +
  transition_time(map_date) +
  ease_aes("linear")

## rendering the animation

animate(anim,
        fps = 10,
        width = 700, 
        height = 500)

## saving the animation

anim_save("tn_drought_animation.gif", anim)
```
