---
title: "US Droughts"
author: "Brittnany Graham"
format: 
  html:
    theme: cosmo
    css: midnight-style.css
    code-fold: true
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

## Background

This dataset was covered by the NY Times and CNN.

The dataset ranges from 2001 to 2021.

```{r}

## Installing tidytuesdayR package and downloading the data

# install.packages("tidytuesdayR")
library(tidytuesdayR)
library(tidyverse)
library(janitor)
library(sf)
library(tigris)

drought <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-07-20/drought.csv')

```

## Cleaning Data

```{r}
## getting a feel for the data
glimpse(drought)

## Clean column names and fix the dates

drought_clean <- drought %>%
  clean_names() %>%
  mutate(
    map_date = ymd(map_date) #converting the dates - be sure to pay attention to the structure of the date
  )
glimpse(drought_clean)


  ## Always look and see what the dataset looks like before you start running the code so this way you can accurately determine what's happening and reduce the amount of headaches

## Clean and Format drought levels

drought_clean <- drought_clean %>%
  mutate(
    drought_lvl = factor(
      drought_lvl,
      levels = c("None", "D0", "D1", "D2", "D3", "D4"),
      labels = c(
        "No Drought",
        "Abnormally Dry",
        "Moderate Drought",
        "Severe Drought",
        "Extreme Drought",
        "Exceptional Drought"
      )
    )
  )

glimpse(drought_clean)
```

## 📍Filter to Tennessee and summarize

Because this is where I live.

```{r}

# Filtering down to Tennessee
tn_drought <- drought_clean %>%
  filter(state_abb == "TN")

## Summarizing

tn_summary <- tn_drought %>%
  group_by(map_date, drought_lvl) %>%
  summarise(
    area_pct = sum(area_pct, na.rm = TRUE),
    .groups = "drop"
  )
```

### Plot

```{r}

## two color palettes for drought
drought_colors <- c(
  "No Drought" = "#FFFFFF",
  "Abnormally Dry" = "#FCD37F",
  "Moderate Drought" = "#FFAA00",
  "Severe Drought" = "#E60000",
  "Extreme Drought" = "#730000",
  "Exceptional Drought" = "#4B0082"
)

midnight_drought <- c(
  "No Drought" = "#0B132B",
  "Abnormally Dry" = "#1C2541",
  "Moderate Drought" = "#3A506B",
  "Severe Drought" = "#5BC0BE",
  "Extreme Drought" = "#C3423F",
  "Exceptional Drought" = "#6A040F"
)

ggplot(tn_summary, aes(x = map_date, y = area_pct, fill = drought_lvl)) +
  geom_area() +
  labs(
    title = "Drought Conditions in Tennessee Over Time",
    x = "Date",
    y = "Percent Area",
    fill = "Drought Level"
  ) +
  scale_fill_manual(values = drought_colors) +
  theme_minimal()
```

## Land versus population impact

I want to look at population level effects of drought in Tennessee.
Since Tennessee's land uses about 41% of it's land for agriculture, that translates to approximately 10.8 million acres of agricultural land.
Various crops and livestock contribute significantly to the state's economy.
Drought inevitably impacts the farmers and surrounding population relying on farmers leading to overall lack of food security in the state.

```{r}

# land versus population impact
tn_compare <- tn_drought %>%
  group_by(map_date, drought_lvl) %>%
  summarise(
    area_pct = sum(area_pct, na.rm = TRUE),
    pop_pct = sum(pop_pct, na.rm = TRUE),
    .groups = "drop"
  )

## pivoting for comparison plotting

tn_compare_long <- tn_compare %>%
  pivot_longer(
    cols = c(area_pct, pop_pct),
    names_to = "metric",
    values_to = "percent"
  )

view(tn_compare_long)
```

### Plotting Area vs. Population

```{r}

ggplot(tn_compare_long, aes(x = map_date, y = percent, fill = drought_lvl)) + 
  geom_area() +
  facet_wrap(~ metric, ncol = 1) +
  scale_fill_manual(values = drought_colors) +
  labs(
    title = "Drought Impact: Land vs. Population in Tennessee",
    y = "Percent",
    x = "Date",
    fill = "Drought Level"
  ) +
  theme_minimal()

```

## Showcasing the Tennessee Map

```{r}

# install.packages("rnaturalearth")
# install.packages("rnaturalearthdata")

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(gganimate)
```

# Tennessee at the county level

I want to start looking at drought conditions at the county level because just one entire state being the same color over time is not...
right

```{r}
# install.packages("tigris")
library(tigris)
```

```{r}
options(tigris_use_cache = TRUE)

tn_counties_sf <- counties(state = "TN", cb = TRUE, class = "sf")
```

## Preparing the drought data at a county scale

```{r}
names(drought_clean)
```

```{r}

drought_tn_county <- read.csv("dm_export_20010101_20211231.csv")

```

## Clean the Tennessee Drought Data

```{r}
## looking at it so we know what we're dealing with

glimpse(drought_tn_county)

```

Now we need to use `mutate()` because we need to:

1.  transform wide columns into long
2.  Adjust the names of the counties
3.  Re-label and re-factor all of the drought levels

```{r}
counties_clean <- drought_tn_county %>%
  clean_names() %>%
  mutate(
    map_date = ymd(map_date) #converting the dates - be sure to pay attention to the structure of the date
  )

## This is where you have to pivot the drought categories from wide to long

counties_long <- counties_clean %>%
  pivot_longer(
    cols = c(d0, d1, d2, d3, d4, none),
    names_to = 'drought_lvl',
    values_to = "percent_area"
  )

## Cleaning and formatting drought levels

counties_long <- counties_long %>%
  mutate(
    county = county %>%
           str_to_upper() %>%
           str_replace("\\s+COUNTY\\s*$", " ") %>%
           str_squish(),
    drought_lvl = factor(
      drought_lvl,
      levels = c("none", "d0", "d1", "d2", "d3", "d4"),
      labels = c(
        "No Drought",
        "Abnormally Dry",
        "Moderate Drought",
        "Severe Drought",
        "Extreme Drought",
        "Exceptional Drought"
      ),
      ordered = TRUE
    )
  )


counties_long <- counties_long %>%
  mutate(
    drought_score = case_when(
      drought_lvl == "No Drought" ~ 0,
      drought_lvl == "Abnormally Dry" ~ 1,
      drought_lvl == "Moderate Drought" ~ 2,
      drought_lvl == "Severe Drought" ~ 3,
      drought_lvl == "Extreme Drought" ~ 4,
      drought_lvl == "Exception Drought" ~ 5
    )
  )

view(counties_long)



```

### Remove zero-area drought rows

```{r}
counties_long <- counties_long %>%
  filter(percent_area > 0)
```

### Keep the MOST SEVERE drought level per county/date

```{r}
tn_drought_map <- counties_long %>%
  group_by(map_date, county) %>%
  slice_max(
    drought_score,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup()

### Verfiy that no duplicates exist, so the return should be '0 rows'

tn_drought_map %>%
  count(map_date, county) %>%
  filter(n > 1)


```

Now we want to reduce to **one** drought level **per county** **per day**

This prevents:

- duplicated counties

- multiple fills per polygon

- broken animations

However, we keep the dominant drought class

```{r}
tn_drought_map <- tn_drought_clean %>%
  group_by(map_date, county) %>%
  slice_max(order_by = drought_score, n = 1, with_ties = FALSE) %>%
  ungroup()

tn_drought_map %>%
  count(drought_lvl)
```

Now we have a completely cleaned dataset

## Next we download Tennessee County geometries

```{r}
options(tigris_use_cache = TRUE)

tn_counties_sf <- counties(
  state = "TN",
  cb = TRUE,
  class = "sf"
)
```

### Clean the county names in the geometry

```{r}
tn_counties_sf <- tn_counties_sf %>%
  mutate(
    NAME = NAME %>%
      str_to_upper() %>%
      str_squish()
  )
```

### Join drought data to county polygons

```{r}

tn_map_sf <- tn_counties_sf %>%
  left_join(
    tn_drought_map,
    by = c("NAME" = "county")
  )
```

**ALWAYS CHECK THE JOIN!!!!!**

```{r}

tn_map_sf %>%
  st_drop_geometry() %>%
  count(drought_lvl)

## if you get NA's the join didn't happen properly
```

------------------------------------------------------------------------

# 🗺️Part 1 - Static Map

------------------------------------------------------------------------

## Filter to one date

```{r}

one_day <- as.Date("2021-07-13")

tn_static <- tn_map_sf %>%
  filter(map_date == one_day)

```

## Create static drought map

```{r}
static <- ggplot(tn_static) +
  geom_sf(aes(fill = drought_lvl),
          color = "black",
          linewidth = 0.15) +
  scale_fill_manual(
    values = drought_colors,
    na.value = "grey85"
  ) +
  labs(
    title = "Tennessee County Drought Conditions",
    subtitle = paste("Date:", one_day),
    fill = "Drought Level"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

ggsave("drought map 2021.jpg", plot = static, height = 5, width = 7)
```

------------------------------------------------------------------------

# Part 2 - Animated County Drought Map

------------------------------------------------------------------------

## Prep the map

```{r}

p <- ggplot(tn_map_sf) +
  geom_sf(aes(fill = drought_lvl,
              group = NAME),
          color = 'black',
          linewidth = 0.1) +
  scale_fill_manual(
    values = drought_colors,
    na.value = "grey85"
  ) +
  labs(
    title = "10 Years of Tennessee County Drought Conditions",
    subtitle = "Date: {frame_time}",
    fill = "Drought Level"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

## Build and render the animation

```{r}

# Building
tn_anim <- p +
  transition_time(map_date) +
  ease_aes("linear")

## Rendering

animate(
  tn_anim,
  fps = 8,
  width = 900,
  height = 650,
  duration = 25
)

## saving the animation

anim_save(
  "tennessee_drought_animation.gif",
  animation = tn_anim
)


## checking the error message

tn_map_sf %>%
  st_drop_geometry() %>%
  count(map_date, NAME) %>%
  filter(n > 1)
```
