---
title: "Tidy Tuesday Week 2"
author: "Brittnany Graham"
format: html
editor: visual
---

> As a publicly-traded company, Amazon releases an annual report every year (with a December 31st year end).
> An annual report is essentially a summary of the company's performance over the past year.
> It includes details on how well the company did financially, what goals were achieved, and what challenges it faced.

```{r}
library(tidyverse)
library(plotly)

report_words_clean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-25/report_words_clean.csv')

```

## Questions to consider:

-   How have the words used change over time?

-   Are there meaningful changes in sentiment from year to year?

-   Which words are likely to appear together in the same annual report?

```{r}
midnight_narrative <- c(
  midnight = "#1D3557",
  coral = "#E76F51",
  blush = "#F4A261",
  rosewood = "#B56576",
  sage = "#6C9A8B",
  linen = "#F1FAEE",
  ink = "#222222"
)
```

## How have the words used changed over time?

```{r}
words <- report_words_clean

head(words)
str(words)
view(words)
```

```{r}
## grouping total words and their counts for the year

word_counts <- words %>%
  group_by(year, word) %>% # group the data by year and word
  summarize(count = n(), .groups = "drop") # counts the number of rows in each group, then ungroups the data frame


# pivot to wide format

heatmap_data <- word_counts %>%
  pivot_wider(names_from = year, values_from = count, values_fill = 0)

# convert to matrix for Plotly

heatmap_matrix <- as.matrix(heatmap_data[ ,-1])
rownames(heatmap_matrix) <- heatmap_data$word

## Creating interactive heatmap

plot_ly(
  x = colnames(heatmap_matrix),
  y = rownames(heatmap_matrix),
  z = heatmap_matrix,
  type = "heatmap",
  colors = c("orange", "red")
 ) %>%
  layout(
    title = "Word Usage Over Years",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Word")
  )

```
