---
title: "Tidy Tuesday Week 2"
author: "Brittnany Graham"
format: 
  html:
    theme: cosmo
    css: midnight-theme.css
    toc: false
    code-fold: true
execute:
  echo: false
  warning: false
  message: false    
editor: visual
---

> As a publicly-traded company, Amazon releases an annual report every year (with a December 31st year end).
> An annual report is essentially a summary of the company's performance over the past year.
> It includes details on how well the company did financially, what goals were achieved, and what challenges it faced.

```{r}
library(tidyverse)
library(plotly)

report_words_clean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-25/report_words_clean.csv')

```

## Questions to consider:

-   How have the words used change over time?

-   Are there meaningful changes in sentiment from year to year?

-   Which words are likely to appear together in the same annual report?

```{r}
midnight_narrative <- c(
  midnight = "#1D3557",
  coral = "#E76F51",
  blush = "#F4A261",
  rosewood = "#B56576",
  sage = "#6C9A8B",
  linen = "#F1FAEE",
  ink = "#222222"
)
```

## How have the words used changed over time?

```{r}
words <- report_words_clean

head(words)
str(words)
view(words)
```

```{r}
## grouping total words and their counts for the year

word_counts <- words %>%
  group_by(year, word) %>% # group the data by year and word
  summarize(count = n(), .groups = "drop") # counts the number of rows in each group, then ungroups the data frame


top_words <- word_counts %>%
  group_by(year) %>%
  slice_max(order_by = count, n = 5) %>% # top 5 words per year
  ungroup()

## capitalizing the words
top_words <- top_words %>%
  mutate(word = str_to_title(word))

### Creating the Stacked Bar Chart

ggplot(top_words, aes(x = factor(year), y = count, fill = word)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top Words Used Each Year",
    x = "Year",
    y = "Word Count",
    fill = "Word"
  ) +
  theme_midnight_narrative()


```

It's clear the company did pretty well at becoming a multi-billion dollar company in its later years.

::: {style="text-align: center; font-size: 0.9em; color: #222222;"}
Tidy Tuesday \| Brittnany Graham \| 2026

Made with RStudio & Quarto
:::
